Tensor denoising and completion based on ordinal observations

### 2020-01-21 ICLM/miaoyan/ordinalT.tex
update draft

@@

\begin{abstract}
Higher-order tensors arise frequently in applications such as neuroimaging, recommendation system, social network analysis, and psychological studies. We consider the problem of low-rank tensor estimation from possibly incomplete, ordinal-valued observations. Two related problems are studied, one on tensor denoising and another on tensor completion. 
%{\color{red}In contrast to the continuous-valued problems, problems with ordinal tensors exhibit an intrinsic phase transition depending on the signal-to-noise ratio. (Not sure whether this should be mentioned)}
We propose a multi-linear cumulative link model, develop a rank-constrained M-estimator, and obtain theoretical accuracy guarantees. 
Our error bound enjoys a faster convergence rate than existing results, and we show that the proposed estimator is minimax optimal under the considered model. 
An efficient completion method is further developed, which guarantees the consistent recovery of an order-$K$ $(d,\ldots,d)$-dimensional tensor using only $O(Kd)$ noisy, quantized observations. We demonstrate the outperformance of our approach over previous methods through applications to the tasks of clustering and collaborative filtering.

@@

\section{Introduction {\color{red}(needs rewriting)}}
Higher-order tensors commonly arise in modern applications. In a context-aware recommendation system, for example, the survey results can be organized as a three-way tensor, where the $(i,j,k)$-th entry indicates the rating from user $i$ to the item $j$ at context $k$. While there are numerous methods for continuous-valued tensors, fewer methods are available for learning tensors based on qualitative observations. Qualitative observations usually take values in a limited set of categories which may be on an ordinal or on a purely nominal scale. The present work focuses on ordinal tensors; namely, the tensor entries are categorical variables with a natural ordering among the categories. 
{\color{red}(Prior work and our contribution. Make a table of comparison.)}

@@

Clearly, the problem (P1) cannot be solved uniformly for all possible $\Theta$. We focus on a class of ``low-rank'' and ``flat'' signal tensors, which is a plausible assumption in practical applications~\cite{zhou2013tensor,bhaskar20151}. Specifically, we consider the parameter space:
\begin{equation}\label{eq:space}
\tP=\left\{\Theta\in\mathbb{R}^{d_1\times \cdots \times d_K} \colon \text{rank}(\Theta)\leq \mr,\ \mnormSize{}{\Theta}\leq \alpha\right\}.
\end{equation}
where $\mr=(r_1,\ldots,r_K)$ denotes the Tucker rank of $\Theta$.

@@

\begin{thm}[Statistical convergence] \label{thm:rate}
Consider an ordinal tensor $\tY\in[L]^{d_1\times\dots\times d_K}$ generated from model~\eqref{eq:model}, with the link function $f$ and the true coefficient tensor $\trueT\in\tP$. Define $r_{\max}=\max_k r_k$. Then, with very high probability, the estimator in~\eqref{eq:estimator} satisfies
\begin{equation}\label{eq:rate}
\text{MSE}(\hat \Theta, \trueT) \leq \min\left( 4\alpha^2,\ {c_2  U^2_\alpha r_{\max}^{K-1}  \over  L^2_\alpha } {\sum_kd_k\over  \prod_k d_k} \right),
\end{equation}
where $c_1, c_2>0$ are two constants that depend only on $K$. 
\end{thm}
Theorem~\ref{thm:rate} provides the statistical converge for the estimator~\eqref{eq:estimator}. Its proof is given in the Supplements. In fact, the proof of this theorem shows that the same statistical rate holds not only for the global optimizer of~\eqref{eq:estimator}, but also for any local optimizer $\check \Theta$ in the level set $\{\check\Theta\in\tP: \logl(\check\Theta)\geq \logl(\trueT)\}$. This suggests that the local optimality itself is not necessarily a severe concern in our context, as long as the objective value at the estimator is large enough. {\color{red}In Section~\ref{algorithm}, we will provide an algorithm-dependent rate quantifying the interplay between the statistical and computational errors.}

@@

We next establish the statistical optimality of our estimator $\hat \Theta$. The result is based on the information theory, and therefore applies to all estimators in $\tP$, including but not limited to $\hat \Theta$ in~\eqref{eq:estimator}. We show that this lower bound nearly matches the upper bound in~\eqref{eq:rate}, as the tensor dimension increases.

@@

\begin{thm}~\label{thm:completion}
Assume the same set-up as in Theorem~\ref{thm:rate}. Suppose that we observe a subset of tensor entries $\{y_\omega\}_{\omega\in\Omega}$, where $\Omega$ is chosen at random with replacement according to a probability distribution $\Pi$. Let $\hat \Theta$ be the solution to~\eqref{eq:estimator}, and assume $r_{\max}=O(1)$. Then, with very high probability over the sampled set $\Omega$, 
\[
\PiFnormSize{}{\Theta-\hat \Theta}^2 \to 0,\quad \text{ as }\quad {|\Omega|\over \sum_k d_k} \ \to \infty. 
\]
\end{thm}
Theorem~\ref{thm:completion} shows that our estimator achieves consistency when sample size is greater than $O(\sum_k d_k)$. {\color{red} add remarks comparing with prior work, and with the degree of freedom}

@@

\section{Experiments}
In this section, we evaluate the empirical performance of our rank-constrained M-estimatation. We consider ordinal tensors with various number of categories, and compare the recovery accuracy with other tensor-based methods. Unless otherwise stated, we generate ordinal-valued tensor under the latent-variable model~\eqref{eq:latent} with i.i.d.\ $N(0,\sigma^2)$ noise. The signal tensor is simulated based on~\eqref{eq:Tucker}, with core tensor generated from i.i.d. N(0,1), and the factor matrices uniformly from orthogonal ensembles. The cut-off points are generated i.i.d.from $N(0,1)$. In each simulation study, we report the summary statistics across $n_{\text{sim}}$ = 30 replications.
\subsection{Finite-sample performance}\label{sec:simulation}

>>>>

\usepackage{amssymb}
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%

@@

\begin{abstract}
Higher-order tensors arise frequently in applications such as neuroimaging, recommendation system, social network analysis, and psychological studies. We consider the problem of low-rank tensor estimation from possibly incomplete, ordinal-valued observations. Two related problems are studied, one on tensor denoising and another on tensor completion. 
%{\color{red}In contrast to the continuous-valued problems, problems with ordinal tensors exhibit an intrinsic phase transition depending on the signal-to-noise ratio. (Not sure whether this should be mentioned)}
We propose a multi-linear cumulative link model, develop a rank-constrained M-estimator, and obtain theoretical accuracy guarantees. 
Our error bound enjoys a faster convergence rate than existing results, and we show that the proposed estimator is minimax optimal under the considered model. 
An efficient completion method is further developed, which guarantees consistent recovery of an order-$K$ $(d,\ldots,d)$-dimensional tensor using only $O(Kd)$ noisy, quantized observations. We demonstrate the outperformance of our approach over previous methods through applications to the tasks of clustering and collaborative filtering.

@@

\section{Introduction {\color{red}(needs rewriting)}}
Higher-order tensors commonly arise in modern applications. In a context-aware recommendation system, for example, the survey results can be organized as a three-way tensor, where the $(i,j,k)$-th entry indicates the rating from user $i$ to the item $j$ at context $k$. While there are numerous methods for continuous-valued tensors, fewer methods are available for learning tensors based on qualitative observations. Qualitative observations usually take values in a limited set of categories which may be on an ordinal or on a purely nominal scale. The present work focuses on ordinal tensors; namely, the tensor entries are categorical variables with a natural ordering among the categories. 
{\color{red}(Prior work and our contribution. See Table~\ref{tab:compare} for comparison.)}

\begin{table*}[http]
\begin{center}
\includegraphics[width=17cm]{compare.pdf}
\end{center}
\vspace{-.5cm}
\caption{Comparison of various low-rank estimation methods based on categorical observations. For an order-$K$ dimensional-$(d,\ldots,d)$ ordinal tensor with $L$-level observations, we report the error bound in the recovered signal (for tensor denoising) and required sample complexity (for tensor completion) as functions of tensor dimension. {\color{red}(font in the table seems different)}}~\label{tab:compare}
\end{table*}

@@

Clearly, the problem (P1) cannot be solved uniformly for all possible $\Theta$. We focus on a class of ``low-rank'' and ``flat'' signal tensors, which is a plausible assumption in practical applications~\cite{zhou2013tensor,bhaskar20151}. Specifically, we consider the parameter space:
\begin{equation}\label{eq:space}
\small \tP=\left\{\Theta\in\mathbb{R}^{d_1\times \cdots \times d_K} \colon \text{rank}(\Theta)\leq \mr,\ \mnormSize{}{\Theta}\leq \alpha\right\}.
\end{equation}
where $\mr=(r_1,\ldots,r_K)$ denotes the Tucker rank of $\Theta$.

@@

\begin{thm}[Statistical convergence] \label{thm:rate}
Consider an ordinal tensor $\tY\in[L]^{d_1\times\dots\times d_K}$ generated from model~\eqref{eq:model}, with the link function $f$ and the true coefficient tensor $\trueT\in\tP$. Define $r_{\max}=\max_k r_k$. Then, with very high probability, the estimator in~\eqref{eq:estimator} satisfies
\begin{equation}\label{eq:rate}
\text{MSE}(\hat \Theta, \trueT) \leq \min\left( 4\alpha^2,\ {c_2  U^2_\alpha r_{\max}^{K-1}  \over  L^2_\alpha } {\sum_kd_k\over  \prod_k d_k} \right),
\end{equation}
where $c_1, c_2>0$ are two constants that depend only on $K$. 
\end{thm}
Theorem~\ref{thm:rate} provides the statistical converge for the estimator~\eqref{eq:estimator}. Its proof is given in the Supplements. In fact, the proof of this theorem shows that the same statistical rate holds not only for the global optimizer of~\eqref{eq:estimator}, but also for any local optimizer $\check \Theta$ in the level set $\{\check\Theta\in\tP: \logl(\check\Theta)\geq \logl(\trueT)\}$. This suggests that the local optimality itself is not necessarily a severe concern in our context, as long as the objective value at the estimator is large enough. In Section~\ref{sec:simulation}, we will perform empirical studies to assess the algorithmic stability and its impact to statistical efficiency.

@@

We next establish the statistical optimality of our estimator $\hat \Theta$. The result is based on the information theory, and therefore applies to all estimators in $\tP$, including but not limited to $\hat \Theta$ in~\eqref{eq:estimator}. We show that this lower bound matches the upper bound in~\eqref{eq:rate} as a function of tensor dimensions.

@@

\begin{thm}~\label{thm:completion}
Assume the same set-up as in Theorem~\ref{thm:rate}. Suppose that we observe a subset of tensor entries $\{y_\omega\}_{\omega\in\Omega}$, where $\Omega$ is chosen at random with replacement according to a probability distribution $\Pi$. Let $\hat \Theta$ be the solution to~\eqref{eq:estimator}, and assume $r_{\max}=O(1)$. Then, with very high probability over the sampled set $\Omega$, 
\[
\PiFnormSize{}{\Theta-\hat \Theta}^2 \to 0,\quad \text{ as }\quad {|\Omega|\over \sum_k d_k} \ \to \infty. 
\]
\end{thm}
Theorem~\ref{thm:completion} shows that our estimator achieves consistency when sample size is greater than $O(\sum_k d_k)$. {\color{red} add remarks comparing with prior work, and with the degree of freedom.}

@@

\section{Experiments}
In this section, we evaluate the empirical performance of our rank-constrained M-estimatation. We consider ordinal tensors with various number of categories, and compare the recovery accuracy with other tensor-based methods. Unless otherwise stated, we generate ordinal-valued tensor under the latent-variable model~\eqref{eq:latent} with i.i.d.\ $N(0,\sigma^2)$ noise. The signal tensor is simulated based on~\eqref{eq:Tucker}, with core tensor generated from i.i.d. N(0,1), and the factor matrices uniformly from orthogonal ensembles. The cut-off points are generated i.i.d.from $N(0,1)$. In each simulation study, we report the summary statistics across $n_{\text{sim}}$ = 30 replications.

\subsection{Finite-sample performance}\label{sec:simulation}
